\chapter{Preface}

The Thomas Jefferson Intermediate Olympiad in Informatics (TJ IOI) is a high-school computer science competition, to be held on Saturday, May 13, 2017, at TJHSST.  We hope to be able to share our passion for computer science with our peers, in hopes of inspiring our fellow students through exposure to concepts not usually taught in the classroom.  In order to help our participants gain a bit of background before the event, we have created a study guide to help ensure that students are equipped with the tools they need to be successful.

The vast majority of the material in this study guide is covered in the curriculum of AP Computer Science A, and we hope that the material will be useful as a review for those of your students also enrolled in that class. That being said, certain additional sections are not covered; namely, those related to trees, graph theory, and dynamic programming. As a result, these sections will account for \textbf{no more than $\frac{1}{4}$ of the actual points in the competition}.

We'd love to discuss how we can help you use this study guide effectively as a student, or as part of a computer science club. If there are errors or room for improvement, we'd certainly like to know about that too. Please don't hestitate to contact any of the officers individually, or send an email to \href{mailto:tjioiofficers@gmail.com}{tjioiofficers@gmail.com}.

We'd like to thank our faculty sponsor, Mr. Thomas Rudwick, for his tireless efforts in helping us plan the event, along with the countless times we've used his room for lunch meetings. We'd also like to thank Mrs. Nicole Kim for her guidance and wisdom from having sponsored TJ IOI in the past, as well as Mr. Stephen Rose for making sure our planning was airtight. Finally, we'd like to thank Dr. Evan Glazer, our principal, and the entire TJ Computer Science Department, for creating a wonderful environment where we can foster our passion for Computer Science.

We hope that you find this study guide useful, and we look forward to seeing you at TJ IOI 2017!

\noindent\textit{The TJ IOI Officers} \\
